---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo=FALSE, results="hide", warning=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE,results="hide",tidy=TRUE,fig.path="figures/")
```

***

## Loading and preprocessing the data
### 1. Load the data
#### - reads csv into dataframe "df"
```{r loaddata}
# assume original data source is in same folder
df <- read.csv("./activity.csv") 
```

### 2. Process/transform the data
#### - adds a datetime variable by concatenating date and interval
```{r processdata}
# creates datetime field
interval_char <- formatC(df$interval,width=4,format="d",flag="0")
datetime_char <- paste(df$date,interval_char)
df$datetime <- strptime(datetime_char,format="%Y-%m-%d %H%M")
```
#### - resulting dataframe "df"
```{r printprocesseddata,results="markup"}
str(df)
```

***
***

## What is mean total number of steps taken per day?
### 1. Make a histogram of the total number of steps taken each day
#### - aggregates total no. of steps per day
```{r aggregatestepsbyday}
# uses the plyr package
library(plyr)
# aggregate the total number of steps per day (while removing NA values)
steps_per_day <- ddply(df,.(date),summarise,total.steps.per.day=sum(steps,na.rm=TRUE))
```
#### - make a histogram
```{r makehistogramofstepsbyday,results='markup'}
# histogram
plot(steps_per_day$date,steps_per_day$total.steps.per.day,type="n",
     main="Histogram of Total Steps per day", xlab="Date", ylab="Total Steps per day")
lines(steps_per_day$date,steps_per_day$total.steps.per.day,type="h",col="red",lwd="3")
```

### 2. Calculate and report the mean and median total number of steps taken per day
#### - compute mean and median steps per day
```{r computemeanmedianstepsperday,results='markup'}
# aggregate mean/median of total steps per day (while removing NA values)
mean_total_steps_per_day <- mean(steps_per_day$total.steps.per.day, na.rm=TRUE)
median_total_steps_per_day <- median(steps_per_day$total.steps.per.day, na.rm=TRUE)
```
#### - mean steps per day is `r mean_total_steps_per_day`, while median steps per day is `r median_total_steps_per_day`

***
***

## What is the average daily activity pattern?
### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
#### - aggregate mean number of steps across all days by each interval (while removing NA values)
```{r aggregatemeanstepsbyinterval}
# aggregate by mean steps
mean_steps_per_interval <- ddply(df,.(interval),summarise,mean.steps=mean(steps, na.rm=TRUE))
```
#### - plot time series
```{r plotmeanstepsbyinterval}
# plot time series of mean steps by interval
plot(mean_steps_per_interval$interval,mean_steps_per_interval$mean.steps,type="l",
     main="Mean steps per interval", xlab="Interval", ylab="Mean Steps")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
#### - find interval with maximum averaged steps
```{r findmaxsteps}
max_steps <- max(mean_steps_per_interval$mean.steps)
interval_max <- mean_steps_per_interval[which.max(mean_steps_per_interval$mean.steps),1]
```
#### - interval at `r interval_max` contains the maximum number of steps (`r max_steps` steps)
```{r plotmaxstepsbyinterval}
# plot point of max steps in same histogram
plot(mean_steps_per_interval$interval,mean_steps_per_interval$mean.steps,type="l",
     main="Mean steps per interval", xlab="Interval", ylab="Mean Steps")
abline(h=max_steps,col="green",lty=3)
abline(v=interval_max,col="green",lty=3)
```

***
***

## Imputing missing values
### 1. Calculate and report the total number of missing values in the dataset
#### - count number of NA in each column (steps, date, interval)
```{r reporttotalmissing}
number_steps_na <- length(df[is.na(df$steps),1])
number_date_na <- length(df[is.na(df$date),2])
number_interval_na <- length(df[is.na(df$interval),3])
```
#### - there are `r number_steps_na` NA values in "steps" column, while `r number_date_na` NA values in "date" column and `r number_interval_na` NA values in "interval" column

### 2. Strategy to fill in missing values
#### - use mean steps in each interval to impute missing
```{r imputemean}
# merge mean steps per interval
df_new <- merge(df,mean_steps_per_interval,by="interval",all.x=TRUE,sort=FALSE)
# if NA then use mean steps
df_new$new.steps <- ifelse(is.na(df_new$steps),df_new$mean.steps,df_new$steps)
# sort by datetime
df_new <- df_new[order(df_new$datetime),]
```

### 3. Create new dataset with missing data filled in
#### - use imputed steps column to create new dataset
```{r createimputeddata}
df_imputed <- df_new[,c("date","interval","datetime","new.steps")]
rownames(df_imputed) <- seq(length=nrow(df_imputed))
```

### 4. Make a histogram of the total number of steps taken per day (again)
#### - aggregates total no. of steps per day (again)
```{r aggregatestepsbydayagain}
# aggregate the total number of steps per day (with imputed values)
steps_per_day_imputed <- ddply(df_imputed,.(date),summarise,total.steps.per.day.imputed=sum(new.steps,na.rm=TRUE))
```
#### - make a histogram
```{r makehistogramofstepsbydayimputed,results='markup'}
# histogram
plot(steps_per_day_imputed$date,steps_per_day_imputed$total.steps.per.day.imputed,type="n",
     main="Histogram of Total Steps per day", xlab="Date", ylab="Total Steps per day")
lines(steps_per_day_imputed$date,steps_per_day_imputed$total.steps.per.day.imputed,type="h",col="red",lwd="3")
```  

#### - compute mean and median, compared to before imputing
```{r computemeanmedianstepsperdayimputed,results='markup'}
# aggregate mean/median of total steps per day (after impute)
mean_total_steps_per_day_imputed <- mean(steps_per_day_imputed$total.steps.per.day.imputed, na.rm=TRUE)
median_total_steps_per_day_imputed <- median(steps_per_day_imputed$total.steps.per.day.imputed, na.rm=TRUE)
diff_mean <- mean_total_steps_per_day_imputed - mean_total_steps_per_day
diff_median <- median_total_steps_per_day_imputed - median_total_steps_per_day
```
#### - mean steps per day (after impute) is `r mean_total_steps_per_day_imputed`, while median steps per day is `r median_total_steps_per_day_imputed`
#### - compared to mean steps per day (before impute) is `r mean_total_steps_per_day`, while median steps per day is `r median_total_steps_per_day`
#### - mean steps per day increased by `r diff_mean`, while median steps per day increased by `r diff_median`

***
***

## Are there differences in activity patterns between weekdays and weekends?
### 1. Create "type.of.day"" variable to indicate "weekday" or "weekend"
#### - create "type.of.day" variable
``` {r createtypeofday}
df_imputed$day <- weekdays(as.Date(df_imputed$date))
weekend <- c("Saturday","Sunday")
df_imputed$type.of.day <- ifelse(df_imputed$day %in% weekend,"weekend","weekday")
df_imputed$type.of.day <- as.factor(df_imputed$type.of.day)
```
#### - results has variable "type.of.day"
``` {r printresults,results='markup'}
# see sample of results
head(df_imputed[df_imputed$type.of.day=="weekday",],5)
head(df_imputed[df_imputed$type.of.day=="weekend",],5)
```


### 2. Time series plot of 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
#### - aggregate mean number of steps across all days by each interval by weekday/weekend
```{r aggregatemeanstepsbyintervaltypeofday}
# aggregate mean steps by interval by type of days,
mean_steps_per_interval_typeofday <- ddply(df_imputed,.(interval,type.of.day),summarise,mean.steps=mean(new.steps, na.rm=TRUE))
```
#### - panel plot time series
```{r plotmeanstepsbyintervaltypeofday}
# use lattice library to plot
library(lattice)
# plot time series of mean steps by interval, conditioned on type of days
xyplot(mean.steps ~ interval | type.of.day, mean_steps_per_interval_typeofday, layout=(c(1,2)), main="Times series of mean steps by interval", xlab="Interval", ylab="Mean Steps", type="l")

```

***